---
title: "startup"
author: "Margaret Janiczek"
date: "2025-07-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Startup: One Time

## Request UKB RAP Access

## Set up project

## Wait for data to be dispensed


# Working on Data: To start

## Log in to UKB RAP

## Start Posit Workbench

## Clone GitHub repository of files 

## (Optional) Grab list of field names, modify them

This is useful if you want to just download a subset of the data to work with going forward. 

### List files in current project

We'll need to figure out where our files are in the RAP project storage.

```{r}
system("dx ls 'Showcase metadata'")
```

### Load the file using dxFUSE

The first way we can access a file in a project is by prepending a `/mnt/project/` to our file path. This lets us access files in our project. 

```{r}
field_data2 <- read.delim("/mnt/project/Showcase metadata/field.tsv")
```

```{r}
head(field_data2)
```

### Alternative: download file into Workspace Storage

If we want to be explicit in downloading files, we can utilize `dx download` within `system()` to download our files from the DX Toolkit.

Note that we 

```{r}
system("dx download '/Showcase metadata/field.tsv'")
list.files()
```

## Alternative: to get table exporter thing to work??

```{r}
field_dx <- read.csv("app81009_20240119215747.dataset.data_dictionary.csv")

library(tidyverse)

field_dx <- field_dx %>%
  mutate(field_mod = name) %>%
  mutate(field_mod2 = str_split_i(field_mod, "_", 1), 
         instance = str_split_i(field_mod, "_i", 2),
         field_mod3 = str_remove(field_mod2, "p"))

field_dx$instance[is.na(field_dx$instance)] <- 0 

field_dx_select <- field_dx %>%
  filter(instance == 0) %>%
  filter(field_mod3 %in% c("eid", field_data_select$field_id)) %>%
 filter(entity == "participant")

write.csv(field_dx_select, file = "please_work.csv")
```


### Load the file into R

```{r}
field_data <- read.delim("field.tsv")
```

```{r}
head(field_data)
```
## Load in the keys for fields used in analysis on Nightengale

```{r}
library(dplyr)
#load("ukb_keys.Rdata") for all pheno data

load("analysis_variable_keys.Rdata")

field_data_select <- field_data2 %>%
  filter(field_id %in% c(as.numeric(char_key$field.showcase), as.numeric(met_key$field.showcase), as.numeric(med_key$field.showcase)))


length(unique(field_data_select$field_id))
  
```


### Writing our file to the RStudio Workspace

```{r}
write.csv(field_data_select, file="updated_field_data.csv")
list.files()


```

### Saving Our file back into Project Storage

```{r}
system("dx upload updated_field_data.csv")
```


## Use Table Exporter to save a .csv file of the data to your Project space


## Wrap up

### Commit and push any file changes


### Save results back to project folder 


### Terminate RStudio session

Be sure to terminate the Posit workbench or it will just keep running (and costing money!!)

# Working on data: subsequent

## Log in to UKB RAP

## (Optional) Use GitHub repository

## (Optional) Connect to GitHub so you can push file changes

## Load in .csv file of data with selected fields

## Run analysis

## Wrap up

### Commit and push any file changes


### Save results back to project folder 


### Terminate RStudio session

Be sure to terminate the Posit workbench or it will just keep running (and costing money!!)


