---
title: "startup"
author: "Margaret Janiczek"
date: "2025-07-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Startup: One Time

## Request UKB RAP Access

## Set up project

## Wait for data to be dispensed


# Working on Data: To start

## Log in to UKB RAP

## Start Posit Workbench

## Clone GitHub repository of files 

## (Optional) Grab list of field names, modify them

This is useful if you want to just download a subset of the data to work with going forward. 

### List files in current project

We'll need to figure out where our files are in the RAP project storage.

```{r}
system("dx ls 'Showcase metadata'")
```

### Load the file using dxFUSE

The first way we can access a file in a project is by prepending a `/mnt/project/` to our file path. This lets us access files in our project. 

```{r}
field_data2 <- read.delim("/mnt/project/Showcase metadata/field.tsv")
```

```{r}
head(field_data2)
```

### Alternative: download file into Workspace Storage

If we want to be explicit in downloading files, we can utilize `dx download` within `system()` to download our files from the DX Toolkit.

Note that we 

```{r}
system("dx download '/Showcase metadata/field.tsv'")
list.files()
```

### Load the file into R

```{r}
field_data <- read.delim("field.tsv")
```

```{r}
head(field_data)
```


### Writing our file to the RStudio Workspace

```{r}
write.csv(field_data2, file="updated_field_data.csv")
list.files()
```

### Saving Our file back into Project Storage

```{r}
system("dx upload updated_field_data.csv")
```


## Use Table Exporter to save a .csv file of the data to your Project space


## Wrap up

### Commit and push any file changes


### Save results back to project folder 


### Terminate RStudio session

Be sure to terminate the Posit workbench or it will just keep running (and costing money!!)

# Working on data: subsequent

## Log in to UKB RAP

## (Optional) Use GitHub repository

## (Optional) Connect to GitHub so you can push file changes

## Load in .csv file of data with selected fields

## Run analysis

## Wrap up

### Commit and push any file changes


### Save results back to project folder 


### Terminate RStudio session

Be sure to terminate the Posit workbench or it will just keep running (and costing money!!)


